---
title: "RLDM Code S2229889"
author: "Sanne van Dijk"
date: "2024-07-02"
output:
  pdf_document:
     latex_engine: xelatex
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/sannevandijk/Desktop/RL&DM/RLDM")

library('dplyr')
library('ggplot2')
library('psych')
library('skimr')
library('tinylabels')
library('papaja')
library('tidyr')
library('tidyverse')
source('helper_functions.r')

dataset4 <- read.csv("dataset4.csv", sep = ",")
rawdata <- as.data.frame(dataset4)

# colors for the plots
color1 <- "darkseagreen1"
color2 <- "lightpink1"
mycolors1 <- c(color1, color2)
mycolors2 <- c(color2, color1)
```

## Inspecting the raw data
### Overview of rawdata
```{r}
skimr::skim(rawdata)
summary(rawdata)
```
### Sample size
```{r}
describe(rawdata)
N <- length(unique(rawdata$ID))
N # N = 12, and there is a total of 9600 observations
```
### Visual inspection
```{r}
# For ggplot, the condition and correct columns should be coded as factors
rawdata_ggplot <- as.data.frame(dataset4) # In order to not get confused later

# Change condition and correct into factors
rawdata_ggplot$condition <- as.factor(rawdata_ggplot$condition)
rawdata_ggplot$correct <- as.factor(rawdata_ggplot$correct)

# Histogram showing the distribution of response times
rt_hist <- ggplot(rawdata_ggplot, aes(x = rt)) +
  geom_histogram(binwidth = 50, fill = color1, color=color2) +
  labs(title = "Distribution of Response Times",
       x = "Response Time",
       y = "Frequency") +
  theme_apa() +
  xlim(0,2000)
print(rt_hist)
```

## Remove extreme variables
```{r}
# Remove outliers using IQR
  # Calculate Q1 (25th percentile) and Q3 (75th percentile)
Q1 <- quantile(rawdata$rt, 0.25, na.rm = TRUE)
Q3 <- quantile(rawdata$rt, 0.75, na.rm = TRUE)

  # Calculate IQR
IQR <- Q3 - Q1

 # Determine bounds for outliers
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

  # Make rawdata numeric
num_rawdata <- as.data.frame(rawdata)
num_rawdata$correct <- as.numeric(num_rawdata$correct, levels = c("1", "0"))

    # Filter the data to remove outliers
Datanooutl <- num_rawdata[num_rawdata$rt >= lower_bound & num_rawdata$rt <= upper_bound, ]

  # Show histogram of the cleaned distribution
DataNO_hist <- ggplot(Datanooutl, aes(x = rt)) +
  geom_histogram(binwidth = 25, fill = color1, color=color2) +
  labs(title = "Distribution of Response Times (Outliers Removed)",
       x = "Response Time",
       y = "Frequency") +
  theme_apa() +
  xlim(0,1200)
print(DataNO_hist)
```
### Outlier inspection
```{r}
outliers <- anti_join(num_rawdata,Datanooutl)

# Overview
skimr::skim(outliers)
summary(outliers)

# Descriptives per participant
outlierspp <- matrix(nrow = 4, ncol = 12, 
                      dimnames = list(c("N", "M rt", "MD rt", "M accuracy"), 1:12))

for (i in 1:12) {
  outlierspp[1,i] <- sum(outliers$ID == i)
  outlierspp[2,i] <- mean(outliers$rt[outliers$ID == i])
  outlierspp[3,i] <- median(outliers$rt[outliers$ID == i])
  outlierspp[4,i] <- mean(outliers$correct[outliers$ID == i])
}
outlierspp

# Table of participant's descriptives after removing outliers
  # Group data by participant (ID) and calculate descriptives
descrip_perpart <- Datanooutl %>%
  group_by(ID) %>%
  summarise(
    mean_rt = mean(rt, na.rm = TRUE),
    sd_rt = sd(rt, na.rm = TRUE),
    median_rt = median(rt, na.rm = TRUE),
    min_rt = min(rt, na.rm = TRUE),
    max_rt = max(rt, na.rm = TRUE),
    mean_accuracy = mean(correct, na.rm = TRUE),
    trials = n()  # Count the number of trials per participant
  )
  # Print the table of descriptives
print(descrip_perpart)
```

## Data Inspection Cleaned Dataset
```{r}
summary(Datanooutl)
describe(Datanooutl)
skimr::skim(Datanooutl)
```
### Descriptives per participant
```{r}
dataWOdesc <- matrix(nrow = 4, ncol = 12, 
                     dimnames = list(c("N", "M rt", "MD rt", "M accuracy"), 1:12))

for (i in 1:12) {
  dataWOdesc[1,i] <- sum(Datanooutl$ID == i)
  dataWOdesc[2,i] <- mean(Datanooutl$rt[Datanooutl$ID == i])
  dataWOdesc[3,i] <- median(Datanooutl$rt[Datanooutl$ID == i])
  dataWOdesc[4,i] <- mean(Datanooutl$correct[Datanooutl$ID == i])
}
dataWOdesc

```
### Descriptives per conditions x corrections
```{r}
conditions <- list(
  list(name = "correct_0", filter = quo(correct == 0)),
  list(name = "correct_1", filter = quo(correct == 1)),
  list(name = "condition_1", filter = quo(condition == 1)),
  list(name = "condition_2", filter = quo(condition == 2)),
  list(name = "condition_1_correct_0", filter = quo(condition == 1 
                                                    & correct == 0)),
  list(name = "condition_1_correct_1", filter = quo(condition == 1 
                                                    & correct == 1)),
  list(name = "condition_2_correct_0", filter = quo(condition == 2 
                                                    & correct == 0)),
  list(name = "condition_2_correct_1", filter = quo(condition == 2 
                                                    & correct == 1))
)

summarycond <- list()
for (cond in conditions) {
  filtered_data <- Datanooutl %>% filter(!!cond$filter)
  summarycond[[cond$name]] <- summary(filtered_data)
}
summarycond
```

## RT and accuracy differences between conditions
```{r}
## Histogram showing RTs for both correct and incorrect as overlaying distributions
histCorrect <- ggplot(Datanooutl, aes(x = rt, group = factor(correct), 
                                  fill = factor(correct))) +
  geom_histogram(binwidth = 5, alpha = .6, position = "identity") +
  papaja::theme_apa() +
  labs(title = "Response Times by Correctness",
       x = "Response Time",
       y = "Frequency",
       fill = "Accuracy") +
  scale_fill_manual(values = mycolors2,
                    labels = c("Correct", "Incorrect")) +  
  xlim(c(0,1500)) +
  annotate("text", x = 50, y = Inf, label = "a", size = 5, 
           fontface = "bold", vjust = -1.5, hjust = 7) + 
  coord_cartesian(clip = "off")  
histCorrect

## Histogram showing RTs for both condition 1 and condition 2 as overlaying distributions
histCond <- ggplot(Datanooutl, aes(x = rt, group = factor(condition), 
                               fill = factor(condition)) )+
  geom_histogram(binwidth = 5, alpha = .6, position = "identity") + 
  papaja::theme_apa() +
  labs(title = "Response Times by Condition",
       x = "Response Time",
       y = "Frequency",
       fill = "Condition") +
  scale_fill_manual(values = mycolors1,
                    labels = c("English Distractors", "Dutch Distractors")) +  
  xlim(c(0,1500)) +
  annotate("text", x = 50, y = Inf, label = "b", size = 5, 
           fontface = "bold", vjust = -1.5, hjust = 7)+
  coord_cartesian(clip = "off") 
histCond
```
### Plots of accuracy by condition
```{r}
# barplot of accuracy by condition
bar_acc <- ggplot(Datanooutl, aes(x = factor(condition), fill = factor(correct))) +
  geom_bar(position = "fill") +
  papaja::theme_apa() +
  labs(title = "Accuracy by Condition",
       x = "Condition",
       y = "Proportion Correct",
       fill = "Accuracy") +
  scale_x_discrete(labels = c("1" = "English Distractor", "2" = "Dutch Distractor")) +
  scale_fill_manual(values = mycolors1,
                    labels = c("Incorrect", "Correct")) +
  annotate("text", x = 1, y = Inf, label = "a", size = 5, 
           fontface = "bold", vjust = -1.5, hjust = 15)+
  coord_cartesian(xlim = c(1,2), clip = "off") 
bar_acc

# dotplot of RT vs accuracy
dot_RT_Acc <- ggplot(Datanooutl, aes(x = rt, y = factor(correct), color = factor(condition))) +
  geom_point(position = position_jitter(height = 0.1), size = 2, alpha = 0.6) +
  scale_color_manual(name = "Condition", values = mycolors1, 
                     labels = c("English Distractor", "Dutch Distractor")) +
  papaja::theme_apa() +
  labs(x = "Reaction Time", y = "Accuracy", 
       title = "Reaction Time vs. Accuracy by Condition") +
  scale_y_discrete(breaks = c(0, 1), labels = c("Incorrect", "Correct")) +
  xlim(c(0, 1500)) +
  annotate("text", x = 50, y = Inf, label = "b", size = 5, fontface = "bold",
           vjust = -1.5, hjust = 6)+
  coord_cartesian(clip = "off") 
dot_RT_Acc
```

## Paired t-test over RT data by condition
```{r}
  # aggregate data by subject, using the median (due to skewed distribution of response times!)
agg <- aggregate(Datanooutl, by = list(Datanooutl$ID, Datanooutl$condition), FUN = median)
agg = subset(agg, select = c(ID, condition, correct, rt))

  # Change agg into wide format and perform a t-test
agg_wide <- agg %>% pivot_wider(names_from = condition, values_from = rt, names_prefix = 'cond_')
res <- t.test(agg_wide$cond_1, agg_wide$cond_2, paired = TRUE)
print(res)

# Differences of mean accuracy between conditions
  # Aggregate data by subject and condition to calculate the mean accuracy
agg_acc <- Datanooutl %>%
  group_by(ID, condition) %>%
  summarize(mean_acc = mean(correct, na.rm = TRUE), .groups = 'drop')

  # Change agg_acc into wide format for the t-test
agg_acc_wide <- agg_acc %>%
  pivot_wider(names_from = condition, values_from = mean_acc, names_prefix = 'cond_')

  # Perform the paired t-test on mean accuracy
res_acc <- t.test(agg_acc_wide$cond_1, agg_acc_wide$cond_2, paired = TRUE)
print(res_acc)
```

## Modelfitting
### Modelfit for each participant in each condition
```{r}
# Fit model for each participant x condition
modelfit <- data.frame(ID = numeric(),
                       condition = numeric(),
                       s = numeric(),
                       A = numeric(),
                       ter = numeric(),
                       b = numeric(),
                       v1 = numeric())
  # Extract parameters and code them into final data frame per subject and condition loop
for (participant in 1:N) {
  for (condition in 1:2) {
    parameters <- fit_data(Datanooutl[Datanooutl$ID == participant & Datanooutl$condition == condition, ]) 
    modelfit[nrow(modelfit) + 1, ] <- unlist(c(participant, condition, parameters))
  }
}
```

***Paramaters of our DDM from the fitting framework provided in the course:***
's' = *The standard deviation of drift rates*: The variability in drift rates. This parameter does however not an easily interpretable cognitive mapping and therefore does not show significant differences between conditions in the provided datasets.
'A' = *The upper limit of the starting point*: The starting point of the evidence accumulation process, it reflects bias
'ter' = *Non-decision time*: The portion of RT occuring independently of the decision-making process.
'b' = *Threshold*: The distance from 0 to the threshold, reflects cautiousness (speed accuract trade-off). Lower thresholds cause faster responses, but reduce accuracy.
'v1' = *Drift rate*: The difficulty level of the problem, higher drift rates cause faster decision-making (easy problem), while lower drift rates cause slower decision-making (hard problem)

### Parameter comparison
```{r}
# Compare Parameters Between Conditions
  ## BIAS
  # Aggregate bias means and SDs
A_means <- aggregate(A ~ condition, data = modelfit, FUN = mean)
A_sds <- aggregate(A ~ condition, data = modelfit, FUN = sd)
  # Bias t-test
A_t_test <- t.test(modelfit$A ~ modelfit$condition, paired = TRUE)
print(A_t_test)

  ## BOUNDARY THRESHOLD
  # Aggregate boundary threshold means and SDs
b_means <- aggregate(b ~ condition, data = modelfit, FUN = mean)
b_sds <- aggregate(b ~ condition, data = modelfit, FUN = sd)
  # Boundary threshold t-test
b_t_test <- t.test(modelfit$b ~ modelfit$condition, paired = TRUE)
print(b_t_test)

  ## DRIFT RATE
  # Aggregate drift rate means and SDs
v1_means <- aggregate(v1 ~ condition, data = modelfit, FUN = mean)
v1_sds <- aggregate(v1 ~ condition, data = modelfit, FUN = sd)
  # Drift rate t-test
v1_t_test <- t.test(modelfit$v1 ~ modelfit$condition, paired = TRUE)
print(v1_t_test)

  ## DRIFT RATE SD
  # Aggregate drift rate SD means and SDs
s_means <- aggregate(s ~ condition, data = modelfit, FUN = mean)
s_sds <- aggregate(s ~ condition, data = modelfit, FUN = sd)
  # Drift rate SD t-test
s_t_test <- t.test(modelfit$s ~ modelfit$condition, paired = TRUE)
print(s_t_test)

  ## NON-DECISION TIME
  # Aggregate non-decision time means and SDs
ter_means <- aggregate(ter ~ condition, data = modelfit, FUN = mean)
ter_sds <- aggregate(ter ~ condition, data = modelfit, FUN = sd)
  # Non-decision time t-test
ter_t_test <- t.test(modelfit$ter ~ modelfit$condition, paired = TRUE)
print(ter_t_test)
```
### Descriptives parameters per condition
```{r}
resultsc1 <- modelfit %>%
  filter(condition == 1)
skimr::skim(resultsc1)

resultsc2 <- modelfit %>%
  filter(condition == 2)
skimr::skim(resultsc2)
```

## Parameter plots
### Boxplots for all parameters
```{r}
modelfit_long <- modelfit %>%
  pivot_longer(cols = c(s, A, ter, b, v1), 
               names_to = "parameter", values_to = "value")

ggplot(modelfit_long, 
       aes(x = factor(condition), y = value, fill = factor(condition))) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(shape = 21, size = 1, position = position_jitter(0.2)) +
  scale_x_discrete(labels = c("English", "Dutch")) +
  facet_wrap(~ parameter, scales = "free_y",  
             labeller = as_labeller(c("s" = "SD drift rate",
                                      "A" = "Starting point",
                                      "ter" = "Non-decision time",
                                      "b" = "Threshold", 
                                      "v1" = "Drift rate"))) +
  scale_fill_manual(values = mycolors1) +  
  papaja::theme_apa() +
  theme(legend.position = "none") +
  labs(x = "Condition", y = "Value", title = "Parameter Distribution by Condition")
```
### Boxplots and slopes for significant parameters
```{r}
# filter for significant parameters 
significant_parameters <- modelfit_long %>%
  filter(parameter %in% c("A"))

# plot
ggplot(significant_parameters, 
       aes(x = factor(condition), y = value, fill = factor(condition))) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(shape = 21, size = 1, position = position_jitter(0.2)) +
  geom_line(aes(group = ID), color = "slateblue1", alpha = 0.7) +  
  scale_x_discrete(labels = c("English", "Dutch")) +
  facet_wrap(~ parameter, scales = "free_y", 
             labeller = as_labeller(c("s" = "SD drift rate", 
                                      "A" = "Starting point",
                                      "ter" = "Non-decision time",
                                      "b" = "Threshold", 
                                      "v1" = "Drift rate"))) +
  scale_fill_manual(values = mycolors1) +  
  papaja::theme_apa() +
  theme(legend.position = "none") +
  labs(x = "Condition", y = "Value", 
       title = "Significant Parameter Distribution by Condition") 
```
### Correlations of parameters
```{r}
# color palette for degree of correlation
color_palette <- colorRampPalette(c(color1, "slateblue1", color2))(200)

# correlation matrix
cor_matrix <- cor(modelfit[,3:7])

# plot
corrplot::corrplot(cor_matrix, method = "circle", type = "upper", 
                  tl.col = "black", tl.srt = 45, 
                  addCoef.col = "black", number.cex = 0.7, 
                  col = color_palette,
                  title = "Correlation Plot of Parameters",
                  mar = c(0,0,1,0)) 
```
